\documentclass[12pt]{article}

\usepackage{amsmath}

\title{Problem Set 2} % Title of the document
\date{\textbf{ECON 6100, Applied Bayesian Methods} \\ March 16th, 2025} % Course name and date
\author{Abdulah Abrahim} % Your name

\begin{document}

\maketitle

\pagestyle{myheadings} % Use the "myheadings" page style
\markright{Abdulah Abrahim} % Set your name in the header

\pagebreak

\begin{homeworkProblem}
\textbf{Problem 1: Use the probability integral transformation method to simulate from the distribution}
\textbf{\begin{gather}
    f(x) = 
    \begin{cases}
        \frac{2}{a^2}x,  & \text{if }0\leq x\leq a \\
        0, & \text{otherwise}
    \end{cases}
\end{gather}
where \( a > 0 \). Set a value for \( a \), simulate various sample sizes, and compare results to the true distribution.}

\begin{solution}

1. **Find the cumulative distribution function (CDF):**
   The CDF is obtained by integrating the PDF:
   \[
   F(x) = \int_{0}^{x} \frac{2}{a^2} t \, dt = \frac{x^2}{a^2}, \quad \text{for } 0 \leq x \leq a.
   \]

2. **Set the CDF equal to a uniform random variable \( U \):**
   Let \( U \sim \text{Uniform}(0,1) \). Then:
   \[
   F(X) = U \implies \frac{X^2}{a^2} = U.
   \]

3. **Solve for \( X \):**
   \[
   X = a \sqrt{U}.
   \]

4. **Simulate samples for various sample sizes:**
   - Choose \( a = 2 \).
   - Generate \( U \) from \( \text{Uniform}(0,1) \).
   - Compute \( X = 2 \sqrt{U} \).

5. 
   - For sample sizes \( n = 100, 1000, 10000 \), F \( F(x) = \frac{x^2}{4} \).
   - As \( n \) increases, the empirical CDF should converge to the true CDF. \textbf{See python code file for histogram visualization of comparison to the true PDF. }
\end{solution}
\end{homeworkProblem}

\pagebreak

\begin{homeworkProblem}
\textbf{Problem 2: Generate samples from the distribution
\begin{gather}
    f(x) = \frac{2}{3}e^{-2x} + 2e^{-3x}
\end{gather}
using the finite mixture approach.}

\begin{solution}
The given density is a finite mixture of two exponential distributions:
\[
f(x) = \frac{2}{3} \cdot 2e^{-2x} + \frac{1}{3} \cdot 3e^{-3x}.
\]
Here, the mixture weights are \( w_1 = \frac{2}{3} \) and \( w_2 = \frac{1}{3} \), and the component distributions are \( \text{Exp}(2) \) and \( \text{Exp}(3) \), respectively.

1. Simulate
   - Generate \( U \sim \text{Uniform}(0,1) \).
   - If \( U \leq \frac{2}{3} \), draw \( X \) from \( \text{Exp}(2) \).
   - Otherwise, draw \( X \) from \( \text{Exp}(3) \).

2. Verify
   - The theoretical mean and variance of the mixture distribution are:
     \[
     \text{Mean} = \frac{2}{3} \cdot \frac{1}{2} + \frac{1}{3} \cdot \frac{1}{3} = \frac{1}{3} + \frac{1}{9} = \frac{4}{9} \approx 0.4444,
     \]
     \[
     \text{Variance} = \frac{2}{3} \cdot \frac{1}{4} + \frac{1}{3} \cdot \frac{1}{9} = \frac{1}{6} + \frac{1}{27} = \frac{11}{54} \approx 0.2037.
     \]

   See python code file for histogram visualization of comparison to the true PDF.. For example, with \( n = 1000 \), the empirical mean should be close to \( 0.4444 \), and the empirical variance should be close to \( 0.2037 \).
\end{solution}
\end{homeworkProblem}

\pagebreak

\begin{homeworkProblem}
\textbf{Problem 3: Draw 500 observations from Beta$(3,3)$ using the accept-reject algorithm. Compute the mean and variance of the sample and compare them to the true values.}

\begin{solution}

1. 
   Use \( \text{Uniform}(0,1) \) as the proposal distribution \( g(x) \).

2. **Find the maximum ratio \( M \):**
   The density of \( \text{Beta}(3,3) \) is:
   \[
   f(x) = \frac{x^{2} (1-x)^{2}}{B(3,3)}, \quad \text{where } B(3,3) = \frac{\Gamma(3)\Gamma(3)}{\Gamma(6)} = \frac{4!}{5!} = \frac{1}{30}.
   \]
   The maximum of \( f(x) \) occurs at \( x = 0.5 \), so:
   \[
   M = \frac{f(0.5)}{g(0.5)} = \frac{(0.5)^2 (0.5)^2 / (1/30)}{1} = \frac{1/16}{1/30} = \frac{30}{16} = 1.875.
   \]

3. Accept-reject algorithm
   - Generate \( U \sim \text{Uniform}(0,1) \) and \( Y \sim \text{Uniform}(0,1) \).
   - Accept \( Y \) if \( U \leq \frac{f(Y)}{M \cdot g(Y)} = \frac{f(Y)}{1.875} \).

4.
   - The theoretical mean and variance of \( \text{Beta}(3,3) \) are:
     \[
     \text{Mean} = \frac{3}{3 + 3} = 0.5,
     \]
     \[
     \text{Variance} = \frac{3 \cdot 3}{(3 + 3)^2 (3 + 3 + 1)} = \frac{9}{6^2 \cdot 7} = \frac{1}{28} \approx 0.0357.
     \]

   See python code file for histogram visualization of comparison to the true PDF. The histogram should closely match the theoretical density, confirming the correctness of the accept-reject algorithm. For example, with \( n = 500 \), the empirical mean should be close to \( 0.5 \), and the empirical variance should be close to \( 0.0357 \).
\end{solution}
\end{homeworkProblem}

\end{document}